SDSS-IV MaNGA: The Radial Profile of Enhanced Star Formation in Close Galaxy Pairs

### 2020-09-08 maintex.tex
edits to the data and samples section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data and Sample}\label{sec:data}

@@

MaNGA is an a IFS survey based at Apache Point Observatory (APO) which used the SDSS (Sloan Digital Sky Survey) 2.5-meter telescope along with two dual-channel BOSS spectrographs \citep{Drory:2015}. MaNGA captures spectra through 17 integral field units (IFUs) with variable numbers of fibers; 19, 37, 61, 91, and 127 fibers covering 12.5\arcsec, 17.5\arcsec, 22.5\arcsec, 27.5\arcsec, and 32.5\arcsec on the sky respectively \citet{Law:2015}. MaNGA is an optical survey with a spectral coverage of 3600$-$10,300 \AA\ with a resolution of R $\sim$ 2000 and a median spectra coverage of 2.5\arcsec\ FWHM \citep{Bundy:2015}.

The MaNGA survey targets galaxies from a subset of 41,154 galaxies from the NASA-Sloan Atlas (NSA v1\_0\_1; \url{http://www.nsatlas.org}) with a redshift range of 0.01 $<$ z $<$ 0.15 and a luminosity range of -17.7 $<$ $\mathcal{M_{\rm i}}$ $<$ -24.0, where $\mathcal{M_{\rm i}}$ is the rest frame i-band magnitude within the survey's elliptical Petrosian apertures. MaNGA plans to cover 10,000 galaxies with a flat stellar mass distribution at two spatial coverages, 1.5 R/\reff\ and 2.5 R/\reff\ (where \reff\ is the radius which contains 50\% of the galaxy's total light). In this work we use the data from the internal data release, MPL-8, which covers 6142 unique galaxies.

@@

Where NUV$-$r is the color and $\mathcal{M}_i$ is the i-band magnitude from the NSA catalog. We have checked the selected star forming galaxies on the BPT diagram \citep{Baldwin:1981} and confirmed that there is not a significant number of AGN in our sample.

@@

We build two different pair samples in this work; the inside-IFU sample which contains paired galaxies where both galaxies are covered by a single MaNGA IFU and the outside-IFU sample where a MaNGA target galaxy is coupled with another galaxies found outside of its MaNGA IFU.

@@

To select paired galaxies out of our object catalog, we inspect the spectra of each object. We use our own {\sc IDL} based spectral fitting code, {\sc SPFIT}, to model the spectra from the MaNGA datacubes. {\sc SPFIT} simultaneously fits emission lines and the stellar continuum with the Levenberg-Marquardt nonlinear least-squares minimization algorithm \citep{Fu:2018}. Emission lines are parameterized as a Gauss-Hermite series and the stellar continuum is the sum of simple stellar populations (SSPs) convolved with the line-of-sight velocity distribution (LOSVD).

The spectra of the identified objects is extracted through a 1\arcsec\ circular aperture and fitted assuming the MaNGA target's redshift and then manually sorted into the following categories; ``good" galaxy spectra, broad-line AGN (active galactic nuclei), foreground star, foreground/background galaxies, or poor S/N objects. The ``good" galaxy spectra are the objects whose spectra is well modeled by {\sc SPFIT} at the target galaxy's redshift, whether it is the target galaxy itself or a nearby companion galaxy. This means that the companion galaxy can be within $\pm$3000 km s$^{-1}$ of the MaNGA target. We found 6573 ``good" objects, 57 broad-line AGN, 836 foreground stars, 319 foreground/background galaxies, and 1546 objects with poor S/N.

From the 6573 galaxies with good spectra, 404 of the MaNGA IFUs have multiple objects within the IFU. We further restrict the sample by setting a relative velocity cut of $\Delta$v $<$ 500 km s$^{-1}$ to remove projected companions. Again, the galaxy sample is also restricted to galaxies with a stellar mass range of \logm\ $=$ 9.0$-$11.5 and galaxies in the Primary or Secondary MaNGA subsamples. We also require that the galaxies are classified as star forming by the CMD diagram. These requirements reduce our sample down to 87 star forming galaxies (SFGs).

@@

We supplement the inside-IFU sample with a set of pairs identified outside of the field of view of the MaNGA IFU. We select these outside-IFU pairs from the NSA catalog. We search for these external pairs by selecting objects with a projected separation from the MaNGA targets of r$_p$ $<$ 50 kpc using the MaNGA target's redshift. We further use a relative velocity cut of $\Delta$v $<$ 500 km s$^{-1}$ to remove projected galaxies from the selection.

@@

The SFR in galaxies has been shown to be dependent on both stellar mass and redshift of the galaxies \citep{Noeske:2007}. This effect needs to be controlled for when comparing the paired and control galaxies to isolate the changes in the sSFR due to merger induced effects.

We build a sample of isolated control galaxies from the remaining MaNGA galaxies by selecting the MaNGA target galaxies which have no spectroscopic companions within r$_p$ $<$ 50 kpc and $\Delta$v $<$ 500 km s$^{-1}$ either inside or outside of the IFU. This gives us a set of 1891 star forming control galaxies to work with.

We then set two separate methods for comparing the paired galaxies with the control galaxies. In the first method, we simply split the galaxies into separate stellar mass bins. We then compare paired galaxies and control galaxies within each stellar mass bin.

In the second method, we control for both stellar mass and redshift. To do this, we set a 0.1 dex stellar mass limit between a paired galaxy and the potential control galaxies. Then we restrict the redshift between the control and paired galaxies, by selecting control galaxies which are drawn from the same MaNGA subsample (e.g. Primary or Secondary) as the paired galaxy. The MaNGA sample has a tight ``banana" shaped distribution in stellar mass and redshift. By setting a 0.1 dex stellar mass cut and requiring that the galaxies be in the same MaNGA subsample, we effectively set a z $\le$ $\sim$0.025 redshift cut.

Setting the above requirements, most paired galaxies will find between 20$-$100 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we will downselect the total number of acquired control galaxies down to 20 controls using a random number generator. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex until at least 20 control galaxies are found.

A small set of paired required extra iterations to find 20 controls; 8 paired galaxies need an extra iteration, 1 paired galaxy need two extra iterations, and 1 paired galaxy needed three iterations.

While the first method is very basic in comparison to the second method, we will see in Section \ref{sec:results} that the first method is sufficient to see the central enhancement to the star formation rate and the mass independence of the star formation enhancement.

@@

We also look at the level of sSFR enhancement as a function of the projected separation between the two galaxies in the bottom row of Figure \ref{fig:ssfr_dmsep}. The $\Delta$log(sSFR) profiles show a clear gradient with the projected separation where close galaxy pairs show higher levels of sSFR enhancement compared to pairs with wide separations. The profiles in the galaxy pairs in the separation range, r$_p$ $=$ 0.0$-$12.5, are generally $\sim$0.1 dex above the median and the the profiles of the galaxy pairs in the separation range, r$_p$ $=$ 37.5$-$50.0, are generally $\sim$0.1 dex below the median.

@@

For galaxies with separations of r$_p$ $<$ 5 kpc and r$_p$ $=$ 30$-$35 kpc we find a $\Delta$log(sSFR) of zero. For the r$_p$ $<$ 5 kpc galaxies, about half of the pairs show signs of tidal features while the others do not. The pairs without the tidal features may be approaching their first pericenter, meaning that the merger-induced gas-inflows have not been triggered yet. The dips in $\Delta$log(sSFR) may also just be due to statistical error from the relatively limited sample size.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data and Samples}\label{sec:data}

@@

MaNGA is an IFS survey at the Apache Point Observatory (APO) which used the SDSS (Sloan Digital Sky Survey) 2.5-meter telescope along with two dual-channel BOSS spectrographs \citep{Drory:2015}. MaNGA captures spectra through 17 integral field units (IFUs) with variable numbers of fibers; 19, 37, 61, 91, and 127 fibers covering 12.5\arcsec, 17.5\arcsec, 22.5\arcsec, 27.5\arcsec, and 32.5\arcsec on the sky respectively \citep{Law:2015}. MaNGA is an optical survey with a spectral coverage of 3600$-$10,300 \AA\ with a resolution of R $\sim$ 2000 and a PSF of 2.5\arcsec\ FWHM \citep{Bundy:2015}.

The MaNGA survey targets galaxies from a subset of 41,154 galaxies from the NASA-Sloan Atlas (NSA v1\_0\_1; \url{http://www.nsatlas.org}) with a redshift range of 0.01 $< z <$ 0.15 and a luminosity range of -17.7 $<$ $\mathcal{M_{\rm i}}$ $<$ -24.0, where $\mathcal{M_{\rm i}}$ is the rest frame i-band magnitude within the survey's elliptical Petrosian apertures. MaNGA plans to cover 10,000 galaxies with a flat stellar mass distribution at two spatial coverages, 1.5 $R$/\reff\ and 2.5 $R$/\reff\ (where \reff\ is the radius which contains 50\% of the galaxy's total light). In this work we use the data from the internal data release, MPL-8, which covers 6142 unique galaxies.

@@

Where $NUV-r$ is the color from SDSS's k-corrected absolute magnitude and $\mathcal{M}_i$ is the i-band magnitude from the NSA catalog. We have checked the selected star forming galaxies on the BPT diagram \citep{Baldwin:1981} to confirm that AGN (active galactic nuclei) will have a minimal impact on our SFG sample.

@@

We build two different pair samples in this work; the inside-IFU sample which contains paired galaxies where both galaxies are covered by a single MaNGA IFU and the outside-IFU sample where a MaNGA target galaxy is coupled with another galaxy found outside of its MaNGA IFU.

@@

To select paired galaxies out of our object catalog, we inspect the spectra of each object. We use our own {\sc idl} based spectral fitting code, {\sc spfit}, to model the spectra from the MaNGA datacubes. {\sc spfit} simultaneously fits emission lines and the stellar continuum with the Levenberg-Marquardt nonlinear least-squares minimization algorithm \citep{Fu:2018}. Emission lines are parameterized as a Gauss-Hermite series and the stellar continuum is the sum of simple stellar populations (SSPs) convolved with the line-of-sight velocity distribution (LOSVD).

The spectra of the identified objects is extracted through a 1\arcsec\ circular aperture and fitted assuming the MaNGA target's redshift and then manually sorted into the following categories; ``good" galaxy spectra, broad-line AGN, foreground star, foreground/background galaxies, or poor S/N objects. The ``good" galaxy spectra are the objects whose spectra are well modeled by {\sc spfit} at the target galaxy's redshift, whether it is the target galaxy itself or a nearby companion galaxy. This means that the companion galaxy can be within $\pm$3000 km s$^{-1}$ of the MaNGA target. We found 6573 ``good" objects, 57 broad-line AGN, 836 foreground stars, 319 foreground/background galaxies, and 1546 objects with poor S/N.

From the 6573 galaxies with good spectra, 404 of the MaNGA IFUs have multiple objects within the IFU. We further restrict the sample by setting a relative velocity cut of $\Delta v$ $<$ 500 km s$^{-1}$ to remove projected companions. Given redshift range of the MaNGA sample and the size of MaNGA's IFUs, the maximum projected separation for a companion galaxy in the IFU is $\sim$40 kpc. Again, the galaxy sample is also restricted to galaxies with a stellar mass range of \logm\ $=$ 9.0$-$11.5 and galaxies in the Primary or Secondary MaNGA subsamples. We also require that the galaxies are classified as star forming by the CMD diagram. These requirements reduce our sample down to 87 star forming galaxies (SFGs).

@@

We supplement the inside-IFU sample with a set of pairs identified outside of the field of view of the MaNGA IFU. We select these outside-IFU pairs from the NSA catalog. We search for these external pairs by selecting objects with a projected separation from the MaNGA targets of $r_p$ $<$ 50 kpc using the MaNGA target's redshift. We further use a relative velocity cut of $\Delta v$ $<$ 500 km s$^{-1}$ to remove projected galaxies from the selection.

@@

To show how the population of galaxy pairs differs from isolated galaxies, we will compare our pair sample to a sample of isolated galaxies in the MaNGA survey. We build this control sample from the MaNGA target galaxies which have no spectroscopic companions within $r_p$ $<$ 50 kpc and $\Delta v$ $<$ 500 km s$^{-1}$ either inside or outside of the IFU. This gives us a control sample of 1891 star forming control galaxies.

It has been shown that the SFR in galaxies is dependent on both the stellar mass and the redshift of the galaxies \citep{Noeske:2007}. We will compare our interacting galaxies with isolated galaxies of a similar stellar mass and redshift, to account for these other dependencies. To do this, we will use two different methods of pair - control comparison.

In the first method, we only control for the stellar mass between the pairs and controls. We split both the pair and control samples into five evenly spaced stellar mass bins over the range, \logm\ $=$ 9.0$-$11.5. The pairs are then compared to their respective controls within each stellar mass bins. While the redshift in not constrained in this method, we will show in Section \ref{sec:mass-bin} that this method is sufficient to reveal the merger induced star formation.

In the second method, we wish to account for both stellar mass and redshift. To do this we select a subsample of 20 control galaxies for each paired galaxy. We first select all of the control galaxies which are within a 0.1 dex stellar mass limit and within the same MaNGA subsample (e.g. Primary or Secondary) as the paired galaxy. Since the MaNGA sample has a tight distribution in stellar mass and redshift, the stellar mass and MaNGA subsample restriction effectively sets a $z$ $\lesssim$ $\sim$0.025 redshift cut.

Setting the above requirements, most paired galaxies will find between 20$-$100 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we will down-select the total number of acquired control galaxies down to 20 controls randomly. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex until at least 20 control galaxies are found. A small set of paired galaxies required extra iterations to find 20 controls; 8 needed an extra iteration, 1 needed two extra iterations, and 1 needed three iterations.

While the first method is less sophisticated in comparison to the second method, we will see in Section \ref{sec:results} that the first method is sufficient to see the central enhancement to the star formation rate and the mass independence of the star formation enhancement.

@@

We also look at the level of sSFR enhancement as a function of the projected separation between the two galaxies in the bottom row of Figure \ref{fig:ssfr_dmsep}. The $\Delta$log(sSFR) profiles show a clear gradient with the projected separation where close galaxy pairs show higher levels of sSFR enhancement compared to pairs with wide separations. The profiles in the galaxy pairs in the separation range, $r_p$ $=$ 0.0$-$12.5, are generally $\sim$0.1 dex above the median and the the profiles of the galaxy pairs in the separation range, $r_p$ $=$ 37.5$-$50.0, are generally $\sim$0.1 dex below the median.

@@

For galaxies with separations of $r_p$ $<$ 5 kpc and $r_p$ $=$ 30$-$35 kpc we find a $\Delta$log(sSFR) of zero. For the $r_p$ $<$ 5 kpc galaxies, about half of the pairs show signs of tidal features while the others do not. The pairs without the tidal features may be approaching their first pericenter, meaning that the merger-induced gas-inflows have not been triggered yet. The dips in $\Delta$log(sSFR) may also just be due to statistical error from the relatively limited sample size.
